#Lab_04 {lab_04}

rob wells, 8/14/2021

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load libraries and establish settings
```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse here
library(tidyverse)
# Load janitor here
library(janitor)


```

### Load data

In this data set, each row represents a city or town in Maryland with the 2010 and 2020 figures for: 

-  Total population 
-  White only population
-  Black or African American only population
-  Hispanic only population 
-  Native American only population
-  Asian only population
-  Pacific Islander (including native Hawiian) only population

Background on this data from [Social Explorer](https://www.socialexplorer.com/data/ACS2016_5yr/metadata/?ds=ACS16_5yr&table=B02001) 
```{r}
# Write code to load the Maryland race by city, place for 2010 - 2020 and call it md_race.
# The data is called lab_04.csv, and is in the same folder as lab_04.qmd (this data notebook)
md_race<- read_csv("lab_04.csv") %>% 
  as.data.frame()

```

## Answer questions

Q1: Name the five cities that had the largest Hispanic population as a percentage of the entire population in 2020 and call it top5_hispanic. Produce a top 5 table for each, selecting the place, 2020 Hispanic, 2020 total and your new calculation.

A1:
```{r}

top5_hispanic <- md_race %>% 
  select(place, x2020_hispanic, x2020_total) %>% 
  mutate(percent_hispanic = (x2020_hispanic/x2020_total)) %>% 
  slice_max(percent_hispanic, n=5) 

top5_hispanic

```

Q2: Compare the top 10 cities with the largest Black population as a percentage of the overall population in 2010 and 2020. Filter for places with more than 20,000 population in 2010. Call your results pct_black. Compare results for 2020 vs 2010. What is different? What surprised you? 

A2: I was surprised that nine out of the 10 cities with the highest Black population had a more than 50% Black population, but the last city on the list, Columbia, only had about a 25% Black population. It's interesting that one of the top 10 cities with Black residence doesn't have that high of a Black population.

```{r}

pct_black <- md_race %>% 
  select(place, x2010_black, x2020_black, x2010_total, x2020_total) %>% 
  mutate(percent_black2010 = (x2010_black/x2010_total)) %>% 
  mutate(percent_black2020 = (x2020_black/x2020_total)) %>% 
  filter(x2010_black > 20000) %>% 
  slice_max(percent_black2020, n=10)

pct_black

```

Q3: Find the 15 communities that had the great gains in Hispanic population from 2010 to 2020 as measured by percentage change. Also calculate percentage change for the community's overall growth during the same period. Filter your results for communities with more than 10,000 people as of 2020. Your results will be hisp_growth. Look at your results, craft a one-paragraph story pitch for your editor. 

A3:
Four Maryland cities, Overlea, Linthicum, Jessup and Glassmanor, all had negative growth for the total population between 2010 and 2020. However, these cities also had tremendous growth in the Hispanic population for all of these cities. The Hispanic population more than tripled in Jessup and Glassmanor, more than quadrupled in Linthicum and more than quintupled in Overlea. What is attracting so many Hispanic people to these cities while more residents are leaving overall?

```{r}

hisp_growth <- md_race %>% 
  select(place, x2010_hispanic, x2020_hispanic, x2010_total, x2020_total) %>% 
  mutate(per_change_hisp = (x2020_hispanic-x2010_hispanic)/x2010_hispanic) %>% 
  mutate(per_change_total = (x2020_total-x2010_total)/x2010_total) %>%
  filter(x2020_total > 10000) %>% 
  slice_max(per_change_hisp, n=15)

hisp_growth


```

