#Lab_03 {lab_03}

rob wells, 8/14/2021

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load libraries and establish settings
```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse here
library(tidyverse)
# Load janitor here
library(janitor)

```

### Load data

In this data set, each row represents a city or town in Maryland with the 2010 and 2020 figures for: 

-  Total population 
-  White only population
-  Black or African American only population
-  Hispanic only population 
-  Native American only population
-  Asian only population
-  Pacific Islander (including native Hawiian) only population

Background on this data from [Social Explorer](https://www.socialexplorer.com/data/ACS2016_5yr/metadata/?ds=ACS16_5yr&table=B02001) 
```{r}
# Write code to load the Maryland race by city, place for 2010 - 2020 and call it md_race.
# The data is called lab_03.csv, and is in the same folder as lab_03.qmd (this data notebook)

md_race <- rio::import("lab_03.csv")

```


## Answer questions

Q1: Which five cities have the largest overall Hispanic population in 2020? Name the five cities that had the largest Hispanic population in 2020. 

A1:
```{r}
# top5_hispanic2020 <- md_race %>% 
#   select(place, x2020_hispanic) %>% 
#   slice_max(x2020_hispanic, n = 5)

md_race %>% 
  select(place, x2020_hispanic) %>% 
  slice_max(x2020_hispanic, n = 5)
```

Q2: Find the cities with the largest Black population and arrange the results in descending order. 

A2:

```{r}
# top_black <- md_race %>% 
#   select(place, x2020_black) %>% 
#   slice_max(x2020_black, n = 100)

top_black <- md_race %>% 
  select(place, x2020_black) %>% 
  arrange(desc(x2020_black))

top_black
```

Q3: Create a table called top_hispanic that shows the top 15 communities with the largest Hispanic population in 2020, and also display the 2010 Hispanic population and the total community population for 2020. Look at your results - what stands out to you? Craft a one-paragraph story pitch for your editor. 

A3:

```{r}
top_hispanic <- md_race %>%
  select(place, x2020_hispanic, x2010_hispanic, x2020_total) %>%
  # arrange(desc(x2020_hispanic)) %>%  not necessary because slice_max descends automatically
  slice_max(x2020_hispanic, n = 15)

```

Some of the cities with the highest Hispanic population are very close to each other. For example, Silver Spring, Langley Park and Adelphi are neighboring communities with East Riverdale, Aspen Hill, Rockville, Germantown, Gaithersburg and Montgomery Village not far away. Most of the top 15 cities with the highest Hispanic population are all Washington, D.C., suburbs. What draws the Hispanic population to this area more than other Maryland metropolitan centers, such as Baltimore or Annapolis? Why did they choose this area and what keeps them here?