#lab_08
Sean Mussenden, Aug 16, 2022

## About this lab

To complete this lab, you need to:
* write code in empty codeblocks provided to answer questions included (look for **Q**).
* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Load libraries and establish settings
**Task**: Load janitor and the tidyverse and sf
```{r}
library(tidyverse)
library(janitor)
library(sf)

```

## Load data

You have been given two spatial dataframes and one regular dataframe to load.  The dataframes are:

* baltimore_homicides.rds | spatial | location and details about each Baltimore homicide between 2007 and 2017, including information about the status (open, closed with arrest, closed without arrest) and the exact location. Details on the data: https://github.com/washingtonpost/data-homicides
* baltimore_tract_shapefiles.rds | spatial | polygons of each census tract in Baltimore
* baltimore_tract_demographics.rds | regular | the total population and white population of each baltimore Census tract in 2017.

**Task** Load the data
```{r}
homicides <- read_rds("data/baltimore_homicides.rds")
shapefiles <- read_rds("data/baltimore_tract_shapefiles.rds")
demographics <- read_rds("data/baltimore_tract_demographics.rds")

```

## Questions

**Q1** Are homicides that occur in majority white Census tracts in Baltimore solved at a higher rate than homicides in non-majority white Baltimore census tracts?
* Write up your findings in three paragraphs that could be dropped into a news story, with the answer to the question as the lede. 
* Include any other relevant details.
* Output a table with relevant results. 
* Output at least one map visualization pertinent to your finding. 

Homicides in majority white Baltimore neighborhoods are solved at a higher rate than in non-majority white neighborhoods. Only 46% of homicides in white neighborhoods are left unsolved, compared to 66% in non-white neighborhoods.

Homicides occur much more frequently in non-white neighborhoods than white neighborhoods. More than 2,500 homicides occurred in non-white neighborhoods in 2017 compared to less than 250 in white neighborhoods. That's 10 times more homicides in non-white neighborhoods. 

There are also more than 15 times as many unsolved homicides in non-white neighborhoods than white neighborhoods. The 1,700 unsolved homicides in non-white neighborhoods greatly outpaces the 100 unsolved homicides in white neighborhoods in 2017. 

**A1** 
```{r}
white_majority <- demographics %>% 
  mutate(pct_white = white_pop/total_pop) %>%
  select(geoid, name, pct_white) %>% 
  mutate(neighborhood_type = case_when(
    pct_white > .50 ~ "majority_white",
    TRUE ~ "minority_white"
  ))

white_majority2 <- white_majority %>%
  inner_join(shapefiles) %>%
  st_as_sf()

murders <- homicides %>%
  st_join(white_majority2) %>%
  as_tibble()

case_status <- murders %>%
  group_by(neighborhood_type, disposition) %>%
  count() %>%
  pivot_wider(names_from=disposition, values_from=n) %>%
  clean_names() %>%
  mutate(total_murders = closed_by_arrest+closed_without_arrest+open_no_arrest) %>%
  mutate(unsolved = open_no_arrest + closed_without_arrest) %>%
  mutate(pct_unsolved = unsolved/total_murders*100)

ggplot() +
  geom_sf(data=shapefiles) +
  geom_sf(data=homicides)

```
