---
title: "Baltimore_Food_Desert"
output: html_notebook
---

Natalie Adams and Matthew Haraden

For our final project, we plan to focus on the prevalence of grocery stores in Baltimore City. Food deserts are an issue in cities, especially in low income and minority areas. We want to look at how food deserts impact Baltimore.

Looking into grocery store data in Baltimore can help us determine if these stores are more accessible to people of a higher income versus people of a lower income. It can also help us see if there are any particular grocery stores that appear more often in higher income communities. The same goes for majority white and non-white majority neighborhoods. This data can be used to find possible discrepancies for grocery store locations in Baltimore.

Questions:
How many grocery stores are in low income neighborhoods?
How many grocery stores are in high income neighborhoods?
How many grocery stores are in majority non-white neighborhoods?
How many grocery stores are in majority white neighborhoods?
How does this compare to convenience stores?

Things to Consider:
How to define high and low income
How to define majority white and non-majority white

Grocery Datasets: 
Supermarkets, Small Grocery Stores and Convenience Stores in Maryland
https://data-clf.hub.arcgis.com/datasets/cce1f648db4b41acafdd66e45d8ad529_595/explore?location=38.803090%2C-77.248650%2C8.00&showTable=true 

Income Datasets:
https://vital-signs-bniajfi.hub.arcgis.com/datasets/bniajfi::median-household-income-community-statistical-area/explore?location=39.284832%2C-76.620524%2C12.52&showTable=true

Race Datasets:
https://vital-signs-bniajfi.hub.arcgis.com/datasets/bniajfi::total-population/explore?layer=0&location=39.284832%2C-76.620524%2C12.52&showTable=true
https://vital-signs-bniajfi.hub.arcgis.com/datasets/bniajfi::percent-of-residents-white-caucasian-non-hispanic-community-statistical-area-1/explore?location=39.284832%2C-76.620524%2C12.52
https://vital-signs-bniajfi.hub.arcgis.com/datasets/bniajfi::percent-of-residents-black-african-american-non-hispanic-community-statistical-area/explore?location=39.284832%2C-76.620524%2C12.52
​​https://vital-signs-bniajfi.hub.arcgis.com/datasets/bniajfi::percent-of-residents-hispanic-community-statistical-area-1/explore?location=39.284832%2C-76.620524%2C12.52
https://vital-signs-bniajfi.hub.arcgis.com/datasets/percent-of-residents-asian-non-hispanic-community-statistical-area-1/explore?location=39.284832%2C-76.620524%2C12.52&showTable=true
https://vital-signs-bniajfi.hub.arcgis.com/datasets/bniajfi::percent-of-residents-all-other-races-hawaiian-pacific-islander-alaskan-native-american-other-race-non-hispanic-community-statistical-area-1/explore?location=39.284832%2C-76.620524%2C12.52
https://vital-signs-bniajfi.hub.arcgis.com/datasets/bniajfi::percent-of-residents-two-or-more-races-non-hispanic-community-statistical-area-1/explore?location=39.284832%2C-76.620524%2C12.52

#load libraries
```{r}
# turn off scientific notation
options(scipen=999)
# load libraries
library(janitor)
library(tidyverse)
library(sf)
library(leaflet)
library(magrittr)
library(ggplot2)
```

#load data
Grocery store data comes from the Johns Hopkins Center for a Livable Future Open Data Portal and median household income and race data comes from the Baltimore Neighborhood Indicators Alliance Vital Signs Open Data Portal (see links above).

```{r}
# food stores in maryland
stores <- read.csv("food_stores.csv")
# median household income by baltimore neighborhood
income <- read.csv("median_household_income.csv")
# baltimore neighborhoods shapefile
neighborhoods <- st_read("income_shapefiles/median_household_income.shp")
# total population by baltimore neighborhood
total_pop1 <- read.csv("total_population.csv")
# white population by baltimore neighborhood
percent_white1 <- read.csv("percent_white.csv")
# black population by baltimore neighborhood
percent_black1 <- read.csv("percent_black.csv")
# hispanic population by baltimore neighborhood
percent_hispanic1 <- read.csv("percent_hispanic.csv")
# asian population by baltimore neighborhood
percent_asian1 <- read.csv("percent_asian.csv")
# other race population by baltimore neighborhood
percent_other1 <- read.csv("percent_other.csv")
# two or more races population by baltimore neighborhood
percent_more1 <- read.csv("percent_two_or_more.csv")
#baltimore population shapefile
population <- st_read("population_shapefiles/total_population.shp")
```

#clean race data
Cleaned the data to select just the neighborhood and 2020 population and then joined all the races together into one table. We did this because we couldn't locate a single dataframe organized by race and neighborhood, so we combined separate ones and checked to make sure the percentages added up to 100.

```{r}
# total population by baltimore neighborhood clean
total_pop2 <- total_pop1 %>% 
  select(CSA2010, tpop20) %>% 
  clean_names() %>%
  rename(neighborhood = 1, total_pop_2020 = 2)

# white population by baltimore neighborhood clean
percent_white2 <- percent_white1 %>% 
  select(CSA2010, pwhite20) %>% 
  clean_names() %>%
  rename(neighborhood = 1, percent_white = 2)

# black population by baltimore neighborhood clean
percent_black2 <- percent_black1 %>% 
  select(CSA2010, paa20) %>% 
  clean_names() %>%
  rename(neighborhood = 1, percent_black = 2)

# hispanic population by baltimore neighborhood clean
percent_hispanic2 <- percent_hispanic1 %>% 
  select(CSA2010, phisp20) %>% 
  clean_names() %>%
  rename(neighborhood = 1, percent_hispanic = 2)

# asian population by baltimore neighborhood clean
percent_asian2 <- percent_asian1 %>% 
  select(CSA2010, pasi20) %>% 
  clean_names() %>%
  rename(neighborhood = 1, percent_asian = 2)

# other race population by baltimore neighborhood clean
percent_other2 <- percent_other1 %>% 
  select(CSA2010, ppac20) %>% 
  clean_names() %>%
  rename(neighborhood = 1, percent_other = 2)

# two or more races population by baltimore neighborhood clean
percent_more2 <- percent_more1 %>% 
  select(CSA2010, p2more20) %>% 
  clean_names() %>%
  rename(neighborhood = 1, percent_two_or_more = 2)

# total baltimore population by race and baltimore neighborhood clean
baltimore_population <- total_pop2 %>% 
  inner_join(percent_white2, by="neighborhood") %>% 
  inner_join(percent_black2, by="neighborhood") %>% 
  inner_join(percent_hispanic2, by="neighborhood") %>% 
  inner_join(percent_asian2, by="neighborhood") %>% 
  inner_join(percent_other2, by="neighborhood") %>% 
  inner_join(percent_more2, by="neighborhood")
```

#clean grocery store data
Selected relevant data points, filtered it for just the Baltimore City stores and cleaned column names. Made separate dataframes for supermarkets, small grocery stores and convenience stores.

```{r}
# baltimore stores
baltimore_stores <- stores %>% 
  select(Main_Categ, Company_Na, County, Address, ZIP_Code, X, Y) %>% 
  filter(County == "Baltimore City") %>% 
  clean_names() %>%
  rename(store_type = 1, name = 2)

# baltimore supermarkets
supermarkets <- read_csv("food_stores.csv") %>% 
  select(Main_Categ, Company_Na, County, Address, ZIP_Code, X, Y) %>% 
  filter(County == "Baltimore City") %>% 
  clean_names() %>%
  rename(store_type = 1, name = 2) %>%
  filter(store_type == "Supermarket") %>% 
  st_as_sf(coords = c("x", "y"), crs = 4269)

# baltimore small grocery stores
small_grocery <- read_csv("food_stores.csv") %>% 
  select(Main_Categ, Company_Na, County, Address, ZIP_Code, X, Y) %>% 
  filter(County == "Baltimore City") %>% 
  clean_names() %>%
  rename(store_type = 1, name = 2) %>%
  filter(store_type == "Small Grocery/Corner Store") %>% 
  st_as_sf(coords = c("x", "y"), crs = 4269)

# baltimore convenience stores
convenience_stores <- read_csv("food_stores.csv") %>% 
  select(Main_Categ, Company_Na, County, Address, ZIP_Code, X, Y) %>% 
  filter(County == "Baltimore City") %>% 
  clean_names() %>%
  rename(store_type = 1, name = 2) %>%
  filter(store_type == "Convenience Store") %>% 
  st_as_sf(coords = c("x", "y"), crs = 4269)
```

#clean median household income data and neighborhoods shapefile
Selected most recent median household income data and cleaned column names.

The median household income in Baltimore is $52,164, according to the U.S. Census Bureau. This is less than the national median household income of $65,000. We chose to use the Baltimore median household income instead of the national median household income because it is more representative of the city. We defined low income as a median household income below Baltimore's $52,164 median household income and high income as above that number.

```{r}
# Load and clean low income neighborhood shapefile
low_income <- st_read("income_shapefiles/median_household_income.shp") %>% 
  clean_names() %>%
  select(csa2010, mhhi20, geometry) %>% 
  rename(neighborhood = 1, median_income=2) %>%
  mutate(
    income_level = case_when(
        median_income < 52164 ~ "low",
        median_income >= 52164 ~ "high"
      )
  ) %>% 
  filter(income_level == "low") %>%
  st_transform(crs=4269)

# Load and clean high income neighborhood shapefile
high_income <- st_read("income_shapefiles/median_household_income.shp") %>% 
  clean_names() %>%
  select(csa2010, mhhi20, geometry) %>% 
  rename(neighborhood = 1, median_income=2) %>%
  mutate(
    income_level = case_when(
        median_income < 52164 ~ "low",
        median_income >= 52164 ~ "high"
      )
  ) %>% 
  filter(income_level == "high") %>%
  st_transform(crs=4269)
```

#How many supermarkets are in low income neighborhoods?
```{r}
# sort coordinates into low income neighborhoods
low_income_supermarkets <- supermarkets %>%
  st_join(low_income) %>% 
  filter(income_level == "low")
```

#How many supermarkets are in high income neighborhoods?
```{r}
# sort coordinates into high income neighborhoods
high_income_supermarkets <- supermarkets %>%
  st_join(high_income) %>% 
  filter(income_level == "high")
```

#plot supermarkets in baltimore
```{r}
ggplot()+
geom_sf(data=neighborhoods)+
geom_sf(data=low_income_supermarkets %>% st_as_sf(coords = c("x", "y"), crs = 4326), mapping=aes(color=income_level))+
geom_sf(data=high_income_supermarkets %>% st_as_sf(coords = c("x", "y"), crs = 4326), mapping=aes(color=income_level))+
theme_minimal()+
labs(title = "Supermarkets in High and Low Income Baltimore Neighborhoods", 
       subtitle = "Johns Hopkins and Baltimore Neighborhood Indicators Alliance Data, 2022",
       caption = "Graphic by Natalie Adams")
```

#How many small grocery stores are in low income neighborhoods?
```{r}
# sort coordinates into low income neighborhoods
low_income_small_grocery <- small_grocery %>%
  st_join(low_income) %>% 
  filter(income_level == "low")

```

#How many small grocery stores are in high income neighborhoods?
```{r}
# sort coordinates into high income neighborhoods
high_income_small_grocery <- small_grocery %>%
  st_join(high_income) %>% 
  filter(income_level == "high")

```

#plot small grocery stores in baltimore
```{r}
ggplot()+
geom_sf(data=neighborhoods)+
geom_sf(data=low_income_small_grocery %>% st_as_sf(coords = c("x", "y"), crs = 4326), mapping=aes(color=income_level))+
geom_sf(data=high_income_small_grocery %>% st_as_sf(coords = c("x", "y"), crs = 4326), mapping=aes(color=income_level))+
theme_minimal()+
labs(title = "Small Grocery and Corner Stores in High and Low Income Baltimore Neighborhoods", 
       subtitle = "Johns Hopkins and Baltimore Neighborhood Indicators Alliance Data, 2022",
       caption = "Graphic by Natalie Adams")
```

#How many convenience stores are in low income neighborhoods?
```{r}
# sort coordinates into low income neighborhoods
low_income_convenience_stores <- convenience_stores %>%
  st_join(low_income) %>% 
  filter(income_level == "low")

```

#How many convenience stores are in high income neighborhoods?
```{r}
# sort coordinates into high income neighborhoods
high_income_convenience_stores <- convenience_stores %>%
  st_join(high_income) %>% 
  filter(income_level == "high")

```

#plot convenience stores in baltimore
```{r}
ggplot()+
geom_sf(data=neighborhoods)+
geom_sf(data=low_income_convenience_stores %>% st_as_sf(coords = c("x", "y"), crs = 4326), mapping=aes(color=income_level))+
geom_sf(data=high_income_convenience_stores %>% st_as_sf(coords = c("x", "y"), crs = 4326), mapping=aes(color=income_level))+
theme_minimal()+
labs(title = "Convenience Stores in High and Low Income Baltimore Neighborhoods", 
       subtitle = "Johns Hopkins and Baltimore Neighborhood Indicators Alliance Data, 2022",
       caption = "Graphic by Natalie Adams")
```



```{r}
#population status
majority_white <- baltimore_population %>% 
  mutate(
    population_status = case_when(
        percent_white < 50 ~ "non_majority_white",
        percent_white >= 50 ~ "majority_white"
      )
  )
#majority and non majority white tables
 majority_white <- percent_white2 %>%
  filter(population_status == "majority_white") %>%
  inner_join(neighborhoods, by="neighborhood")

 non_majority_white<- percent_white2 %>%
  filter(population_status == "non_majority_white") %>%
  inner_join(neighborhoods, by="neighborhood")
```


